A Bayesian approach to  uncertainty in word embedding bias estimation

### 2023-06-15 paper/paperDraft.Rmd
paper sent to arxiv

---
title: 'Taking uncertainty in word embedding bias estimation seriously: a  bayesian
  approach'
author: "Alicja Dobrzeniecka and Rafal Urbaniak"
output:
  pdf_document:
    number_sections: yes
    df_print: kable
    keep_tex: yes
    includes:
      in_header: Rafal_latex7.sty
  html_document:
    df_print: paged
classoption: dvipsnames,enabledeprecatedfontcommands
fontsize: 12pt
documentclass: scrartcl
urlcolor: blue
bibliography: ../references/cosineReferences1.bib
csl: ../references/transactions-on-computational-logic.csl
---

@@

@@

Bayesian data analysis takes prior probability distributions, a mathematical model structure and the data, and returns the posterior probability distributions over the parameters of interest, thus capturing our uncertainty about their actual values. One important difference between such  a result and the result of a classical statistical analysis is that classical confidence intervals (CIs) have a rather complicated and somewhat confusing interpretation, which has little to do with the posterior probability distribution.^[Here are a few usual problems. CIs are often mistakenly interpreted as providing  the probability that a resulting confidence interval contains the true value of a parameter. CIs bring confusion also with regard to  precision, it is a common mistake to interpret narrow intervals as the ones corresponding to a more precise knowledge. Another fallacy is to associate CIs with likelihood and stating that values within a given interval are more probable than the ones outside it. The theory of confidence intervals does not support the above interpretations. CIs should be plainly interpreted as a result of certain procedure (there are many ways to obtain CIs from a given set of data) that will in the long run contain the true value if the procedure is performed a fixed amount of times. For a nice survey and explanation of these misinterpretations, see  [@Morey2015confidenceFallacy]. For a psychological study of the occurrence of such  misinterpretations, see REF. In this study, 120 researchers and 442 students were asked to assess the truth value of six false statements involving different interpretations of a CI.  Both researchers and students endorsed, on average, more than three of these statements. See also [@Hoekstra2014Misinterpretation] for a list of problems with the classical approach.]

@@

# Loose ends and conclusions
\label{sec:related}

[@DBLP:journals/corr/abs-1811-07253] employ Bayesian methods to estimate uncertainty in NLP tasks, but they apply their Bayesian Neural Networks-based method to sentiment analysis or named entity recognition, not to bias. [@Ethayarajh2020Bernstein] correctly argues that bias estimate should not be expressed as a single number without taking into account that the estimate is made using a sample of data and therefore has intrinsic uncertainty. The authors   suggest to use Bernstein bounds to gauge  the uncertainty in terms of  confidence intervals. We do not discuss this approach extensively, as we think that confidence intervals are quite problematic for several reasons, among others the confusing interpretability.  We do not think that  Bernstein bounds provide the best solution to the problem. Applying this method to a popular WinoBias dataset leads to the conclusion that more than 11903 samples are needed to claim a 95\% confidence interval for a bias estimate. This amount vastly exceeds the existing word lists for bias estimation. We propose a more realistic Bayesian method. Our conclusion is still that the word lists are sometimes too small, but at least allows for gauging uncertainty as we go on to improve our methodology and  extend the lists gradually.

 [@schröder2021evaluating]  criticize some exisisting bias metrics such as \textsf{MAC} or \textsf{WEAT} on the grounds of them not satisfying some general formal principles, such as magnitude-comparability, and they propose a modification, called \textsf{SAME}. [@may-etal-2019-measuring] develop a generalization of \textsf{WEAT} meant to apply to sets of sentences, \textsf{SEAT}, which basically applies \textsf{WEAT} to vector representations of sentences.  The authors, however, still pre-average and play the game of finding a single-number metric, so our remarks  apply.

\todo{More references to similar papers}

A Bayesian data analysis with hierarchical models of cosine distances between protected words, control group words, and stereotypical attributes  provides more modest and realistic assessment of the uncertainty involved. It reveals that much complexity is hidden when one instead chases single bias metrics present in the literature.

After introducing the method, we apply it to multiple word embeddings and results of supposed debiasing, putting forward some general observations that are not exactly in line with the usual picture painted in terms of \textsf{WEAT} or \textsf{MAC} (and the problem generalizes to any approach that focuses on chasing a single numeric metric): the word list sizes and sample sizes used in the studies are usually small.  posterior density intervals are fairly wide,   often the differences between associated, different and neutral human predicates, are not very impressive. Also, a preliminary inspection suggests that  the desirability of changes obtained by the usual debiasing methods is debatable. The tools that we propose, however, allow for a more fine-grained and multi-level evaluation of bias and debiasing in language models without losing modesty about the uncertainties involved.

The short general lesson here is this: it's complicated. Instead of pretending it isn't and chasing single-number metrics, we should rather devote attention to more nuanced and methodologically sound analysis.

>>>>

---
title: 'A Bayesian approach to  uncertainty in word embedding bias estimation'
author: "Alicja Dobrzeniecka and Rafal Urbaniak"
output:
  pdf_document:
    number_sections: yes
    df_print: kable
    keep_tex: yes
    includes:
      in_header: Rafal_latex7.sty
  html_document:
    df_print: paged
classoption: dvipsnames,enabledeprecatedfontcommands
fontsize: 12pt
documentclass: scrartcl
urlcolor: blue
bibliography: ../references/cosineReferences1.bib
csl: ../references/transactions-on-computational-logic.csl
---

@@

\textbf{Abstract.} Multiple measures, such as WEAT or MAC, attempt to quantify the magnitude of  bias present in  word embeddings  in  terms of a single-number  metric. However, such  metrics and the related statistical significance calculations rely on treating pre-averaged data as individual  data points and employing bootstrapping techniques with low sample sizes. We show that similar results can be easily obtained using such methods even if the data are generated by a null model lacking the intended bias. Consequently, we argue that this approach generates false confidence. To address this issue, we propose a Bayesian alternative: hierarchical Bayesian modeling, which enables a  more uncertainty-sensitive inspection of bias in  word embeddings at different levels of granularity. To showcase our method, we apply it to  \emph{Religion, Gender}, and \emph{Race} word lists from the original research,  together with our control neutral word lists. We   deploy the method using  \emph{Google, Glove}, and \emph{Reddit} embeddings. Further, we utilize our approach to evaluate a debiasing technique applied to the \emph{Reddit} word embedding. Our findings reveal a more complex landscape than suggested by the proponents of single-number metrics. The datasets and source code for the paper are publicly available.^[\url{https://github.com/efemeryds/Bayesian-analysis-for-NLP-bias}]

@@

Bayesian data analysis takes prior probability distributions, a mathematical model structure and the data, and returns the posterior probability distributions over the parameters of interest, thus capturing our uncertainty about their actual values. One important difference between such  a result and the result of a classical statistical analysis is that classical confidence intervals (CIs) have a rather complicated and somewhat confusing interpretation, which has little to do with the posterior probability distribution.^[Here are a few usual problems. CIs are often mistakenly interpreted as providing  the probability that a resulting confidence interval contains the true value of a parameter. CIs bring confusion also with regard to  precision, it is a common mistake to interpret narrow intervals as the ones corresponding to a more precise knowledge. Another fallacy is to associate CIs with likelihood and stating that values within a given interval are more probable than the ones outside it. The theory of confidence intervals does not support the above interpretations. CIs should be plainly interpreted as a result of certain procedure (there are many ways to obtain CIs from a given set of data) that will in the long run contain the true value if the procedure is performed a fixed amount of times. For a nice survey and explanation of these misinterpretations, see  [@Morey2015confidenceFallacy]. For a psychological study of the occurrence of such  misinterpretations, see [@Hoekstra2014Misinterpretation]. In this study, 120 researchers and 442 students were asked to assess the truth value of six false statements involving different interpretations of a CI.  Both researchers and students endorsed, on average, more than three of these statements.]

@@

# Related works and conclusions
\label{sec:related}

There are a few related papers, whose discussion goes beyond the scope of this paper:

-  [@DBLP:journals/corr/abs-1811-07253] employ Bayesian methods to estimate uncertainty in NLP tasks, but they apply their Bayesian Neural Networks-based method to sentiment analysis or named entity recognition, not to bias.

- [@Ethayarajh2020Bernstein] correctly argues that a bias estimate should not be expressed as a single number without taking into account that the estimate is made using a sample of data and therefore has intrinsic uncertainty. The authors   suggest using Bernstein bounds to gauge  the uncertainty in terms of  confidence intervals. We do not discuss this approach extensively, as we think that confidence intervals are quite problematic for several reasons, among others the confusing interpretability.  We do not think that  Bernstein bounds provide the best solution to the problem. Applying this method to a popular WinoBias dataset leads to the conclusion that more than 11903 samples are needed to claim a 95\% confidence interval for a bias estimate. This amount vastly exceeds the existing word lists for bias estimation. We propose a more realistic Bayesian method. Our conclusion is still that the word lists are sometimes too small, but at least they allow for gauging uncertainty as we go on to improve our methodology and  extend the lists gradually.

-  [@schröder2021evaluating]  criticize some existing bias metrics such as \textsf{MAC} or \textsf{WEAT} on the grounds of them not satisfying some general formal principles, such as magnitude-comparability, and they propose a modification, called \textsf{SAME}.

- [@may-etal-2019-measuring] develop a generalization of \textsf{WEAT} meant to apply to sets of sentences, \textsf{SEAT}, which basically applies \textsf{WEAT} to vector representations of sentences.  The authors, however, still pre-average and play the game of finding a single-number metric, so our remarks  apply.

- [@Guo2021CEAT] introduce the Contextualized Embedding Association Test Intersectional, meant to apply to dynamic word embeddings and, importantly develop methods for intersectional bias detection. The measure is a generalization of the WEAT method. The authors do inspect a distribution of effect sizes that arises from the consideration of various possible contexts, but they continue to standardize the difference in averaged means and  use a single-number summary: the weighted mean of the effect sizes thus understood. The method, admittedly, deserves further evaluation which goes beyond the scope of this paper.

To summarize, a Bayesian data analysis with hierarchical models of cosine distances between protected words, control group words, and stereotypical attributes  provides more modest and realistic assessment of the uncertainty involved. It reveals that much complexity is hidden when one instead chases single bias metrics present in the literature.

After introducing the method, we applied it to multiple word embeddings and results of supposed debiasing, putting forward some general observations that are not exactly in line with the usual picture painted in terms of \textsf{WEAT} or \textsf{MAC} (and the problem generalizes to any approach that focuses on chasing a single numeric metric): the word list sizes and sample sizes used in the studies are usually small.  Posterior density intervals are fairly wide.  Often the differences between associated, different and neutral human predicates, are not very impressive. Also, a preliminary inspection suggests that  the desirability of changes obtained by the usual debiasing methods is debatable. The tools that we propose, however, allow for a more fine-grained and multi-level evaluation of bias and debiasing in language models without losing modesty about the uncertainties involved.

The short general and somewhat disappointing lesson here is this: things are complicated. Instead of chasing single-number metrics, we should rather devote attention to more nuanced analysis.
